Reformulating phylogenetic mixed models to improve flexibility and speed

### 2020-02-17 phyloglmm_ms.tex
started working on trimming

Phylogenetic comparative methods (PCMs) are a powerful technique to account for evolutionary relatedness when analyzing species traits.
Given a known phylogenetic tree, PCMs explore the relationships among species traits or distributions while taking the underlying relationships of the species into account; they can be used to control statistically for phylogenetic relationships, to quantify phylogenetic signal (a measure of the dependence among species responses due to their evolutionary relationships) in trait distributions, or both.

Unlike standard regression-based statistical models, where all of the predictor variables of interest (for example, species traits and environmental covariates) are directly observable,  PCMs using phylogenetic regression use phylogenetic relationships to incorporate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic, hansen2012interpreting}. 
While a wide range of tools is available for phylogenetic regression, existing procedures may be either insufficiently flexible or too computationally demanding when analyzing large volumes of data.
When faced with these constraints, researchers typically search for ways to simplify their analyses.

@@

Classic phylogenetic regression uses a statistical model in which phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the dependent-variable trait evolves along the branches of the phylogeny according to a Brownian-motion evolutionary model \citep{felsenstein1985phylogenies}. 
If in addition the residuals are normally distributed and observed without any additional error or within-species variation, Felsenstein's method of phylogenetically independent contrasts  \citep[PICS:][]{felsenstein1985phylogenies} is sufficient to account for the phylogenetic correlation.
More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized, housworth2004phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon PICs by considering different (non-Gaussian) response distributions and by accounting for evolutionary processes other than the simple Brownian motion model. 
These methods partition residual variation into two components: (1) uncorrelated, or independent, residual variation (observation error or tip variation) and (2) phylogenetic signal (evolutionary process error) \citep{hansen2012interpreting, housworth2004phylogenetic}.
If each species' traits are observed more than once, possibly under different conditions, we can potentially distinguish a third level of variation; in this case, phylogenetic variation and tip variation can both be considered part of the evolutionary process error (which we will call tip variation or intercept-level variation) while the among species residual variation is associated with among observation variation within each species.
Although many studies include multiple observations per species, phylogenetic analyses rarely take advantage of such information to partition variability more finely.

Classic phylogenetic regressions usually allow the uncertainty in the response (trait or distribution) to evolve according to the phylogenetic relationship across species, but the effects of the predictor variables may evolve according to the phylogenetic relationship across species as well. 
For example, suppose we have examined a collection of species that came from two groups, and wish to know whether their brain size (Y) is proportional to their body size (X) (Felsenstein's (1985) example using a mixed-effect model. 
The standard phylogenetic mixed-effect model allows for phylogenetic correlations in the intercept of the relationship between body and brain size. 
However, species within taxonomic groups with similar body sizes may vary in brain size (random intercepts), or taxonomic groups may vary in the relationship between brain and body size (random slopes). 
Analogous to phylogenetically correlated variation in intercepts, phylogenetic random slopes allow the relationship between predictors and responses to vary between species in a phylogenetically correlated way.
Several recent studies have incorporated some form of phylogenetic variation using random effects in addition to the standard phylogenetic correlations in the intercepts and looked at species response to phylogenetic variation with changes in environmental factors \citep{nowakowski2018phylogenetic, li2017canfun}.
The tools available for extending phylogenetic relationships to predictor variables (or predictor level variation) in a standard frequentist framework are relatively inflexible; thus, biologists needing to fit random-slopes model typically turn to more flexible Bayesian approaches, despite their additional computational burden \citep{hadfield2010mcmc, burkner2018brms}.
Table \ref{table:model} summarizes currently available methods for phylogenetic comparative analysis.

>>>>

Phylogenetic comparative methods (PCMs) are a powerful technique to account for evolutionary relatedness when analyzing species traits.
Given a known phylogeny, PCMs model the relationships among species traits while accounting for relatedness; they can be used to control statistically for phylogeny, to quantify phylogenetic signal in traits (dependence among traits due to evolutionary relationships), or both.

Unlike standard statistical models, where all of the predictor variables of interest (such as species traits and environmental covariates) are directly observable, PCMs use phylogenetic relationships to incorporate the unobserved process of trait evolution \citep{felsenstein1985phylogenies, butler2004phylogenetic, hansen2012interpreting}. 
While a wide range of tools is available for phylogenetic regression, existing procedures may be either insufficiently flexible or too computationally demanding when analyzing large volumes of data.
When faced with these constraints, researchers often simplify their analyses, at the risk of neglecting important processes.

@@

In classic phylogenetic regression, phylogenetic correlation in the residuals from a regression between two species-level traits arises because the residual variation in the response trait evolves along the branches of the phylogeny according to a Brownian-motion evolutionary model \citep{felsenstein1985phylogenies}. 
If the residuals are normally distributed and observed without additional error or within-species variation, Felsenstein's method of phylogenetically independent contrasts  \citep[PICS:][]{felsenstein1985phylogenies} is sufficient to account for the phylogenetic correlation.
More recent approaches --- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized, housworth2004phylogenetic}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- extend PICs by considering different (non-Gaussian) response distributions and by accounting for evolutionary models other than Brownian motion.
These methods partition residual variation into two components: (1) independent residual variation (tip variation, which is confounded with observation error) and (2) phylogenetic signal (evolutionary process error) \citep{hansen2012interpreting, housworth2004phylogenetic}.
If each species' traits are measured multiple times, possibly under different conditions, we can distinguish a third level of variation; in this case, phylogenetic variation and tip variation are both part of the evolutionary process while the observation error can be independently identified from within-species variation.
Although many studies include multiple observations per species, phylogenetic analyses rarely take advantage of such information to partition variability more finely.

Classic phylogenetic regressions allow the residuals to evolve along the phylogeny, but the effects of the predictor variables may evolve along the phylogeny as well; extending PCMs to allow for multiple sources of variation lead to the  
For example, suppose we wish to fit a PCM to predict species' brain size (Y) from their body size (X) \cite{felsenstein1985phylogenies} using a mixed-effect model.

The standard phylogenetic mixed-effect model allows for phylogenetic correlations in the intercept of the relationship between body and brain size. 
However, species within taxonomic groups with similar body sizes may vary in brain size (random intercepts), or taxonomic groups may vary in the relationship between brain and body size (random slopes). 
Analogous to phylogenetically correlated variation in intercepts, phylogenetic random slopes allow the relationship between predictors and responses to vary between species in a phylogenetically correlated way.
Several recent studies have incorporated some form of phylogenetic variation using random effects in addition to the standard phylogenetic correlations in the intercepts and looked at species response to phylogenetic variation with changes in environmental factors \citep{nowakowski2018phylogenetic, li2017canfun}.
The tools available for extending phylogenetic relationships to predictor variables (or predictor level variation) in a standard frequentist framework are relatively inflexible; thus, biologists needing to fit random-slopes model typically turn to more flexible Bayesian approaches, despite their additional computational burden \citep{hadfield2010mcmc, burkner2018brms}.
Table \ref{table:model} summarizes currently available methods for phylogenetic comparative analysis.
